---
title: "Deployment and sandboxing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Deployment and sandboxing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## The sandbox

With `sandbox = TRUE`, the child R process runs under restrictions set by the
operating system. It can read what R needs to run and write to a temp
directory. It can't use the network or touch the rest of your files. Each
platform does this differently, as described below.

### What happens on each call

A sandboxed call goes through five steps:

```{=html}
<figure class="hairline">
<svg role="img" aria-label="Create, configure, execute, collect, clean up; the child can be reused for the next call" viewBox="0 0 920 172" xmlns="http://www.w3.org/2000/svg"><defs>
<marker id="lc-arrow" viewBox="0 0 10 10" refX="9" refY="5" markerWidth="7" markerHeight="7" orient="auto-start-reverse">
<path d="M1 1L9 5L1 9" fill="none" stroke="#2b1f12" stroke-width="1"/></marker>
<pattern id="lc-hatch" width="6" height="6" patternUnits="userSpaceOnUse" patternTransform="rotate(45)">
<line x1="0" y1="0" x2="0" y2="6" stroke="#bf5a36" stroke-width="0.6" opacity="0.55"/></pattern>
</defs><text x="20" y="34" font-family="Space Mono, ui-monospace, monospace" font-size="8.5" fill="#bf5a36" text-anchor="start" font-weight="400" letter-spacing="1.4" >01</text><text x="42" y="34" font-family="Space Mono, ui-monospace, monospace" font-size="8.5" fill="#6b5638" text-anchor="start" font-weight="400" letter-spacing="1.6" >CREATE</text><rect x="20" y="50" width="150" height="70" fill="none" stroke="#2b1f12" stroke-width="0.75"/><text x="95.0" y="82.0" font-family="Space Mono, ui-monospace, monospace" font-size="10.5" fill="#2b1f12" text-anchor="middle" font-weight="700" letter-spacing="1.2" >START R</text><text x="95.0" y="96.0" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="0.4" >a child process</text><text x="200" y="34" font-family="Space Mono, ui-monospace, monospace" font-size="8.5" fill="#bf5a36" text-anchor="start" font-weight="400" letter-spacing="1.4" >02</text><text x="222" y="34" font-family="Space Mono, ui-monospace, monospace" font-size="8.5" fill="#6b5638" text-anchor="start" font-weight="400" letter-spacing="1.6" >CONFIGURE</text><rect x="200" y="50" width="150" height="70" fill="none" stroke="#2b1f12" stroke-width="0.75"/><text x="275.0" y="75.0" font-family="Space Mono, ui-monospace, monospace" font-size="10.5" fill="#2b1f12" text-anchor="middle" font-weight="700" letter-spacing="1.2" >LOCK IT DOWN</text><text x="275.0" y="89.0" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="0.4" >sandbox and</text><text x="275.0" y="102.0" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="0.4" >resource limits</text><path d="M170 85H198" fill="none" stroke="#2b1f12" stroke-width="0.75" marker-end="url(#lc-arrow)"/><text x="380" y="34" font-family="Space Mono, ui-monospace, monospace" font-size="8.5" fill="#bf5a36" text-anchor="start" font-weight="400" letter-spacing="1.4" >03</text><text x="402" y="34" font-family="Space Mono, ui-monospace, monospace" font-size="8.5" fill="#6b5638" text-anchor="start" font-weight="400" letter-spacing="1.6" >EXECUTE</text><rect x="380" y="50" width="150" height="70" fill="none" stroke="#2b1f12" stroke-width="0.75"/><text x="455.0" y="75.0" font-family="Space Mono, ui-monospace, monospace" font-size="10.5" fill="#2b1f12" text-anchor="middle" font-weight="700" letter-spacing="1.2" >RUN THE CODE</text><text x="455.0" y="89.0" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="0.4" >and answer its</text><text x="455.0" y="102.0" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="0.4" >tool calls</text><path d="M350 85H378" fill="none" stroke="#2b1f12" stroke-width="0.75" marker-end="url(#lc-arrow)"/><text x="560" y="34" font-family="Space Mono, ui-monospace, monospace" font-size="8.5" fill="#bf5a36" text-anchor="start" font-weight="400" letter-spacing="1.4" >04</text><text x="582" y="34" font-family="Space Mono, ui-monospace, monospace" font-size="8.5" fill="#6b5638" text-anchor="start" font-weight="400" letter-spacing="1.6" >COLLECT</text><rect x="560" y="50" width="150" height="70" fill="none" stroke="#2b1f12" stroke-width="0.75"/><text x="635.0" y="75.0" font-family="Space Mono, ui-monospace, monospace" font-size="10.5" fill="#2b1f12" text-anchor="middle" font-weight="700" letter-spacing="1.2" >RETURN</text><text x="635.0" y="89.0" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="0.4" >the result and</text><text x="635.0" y="102.0" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="0.4" >the output</text><path d="M530 85H558" fill="none" stroke="#2b1f12" stroke-width="0.75" marker-end="url(#lc-arrow)"/><text x="740" y="34" font-family="Space Mono, ui-monospace, monospace" font-size="8.5" fill="#bf5a36" text-anchor="start" font-weight="400" letter-spacing="1.4" >05</text><text x="762" y="34" font-family="Space Mono, ui-monospace, monospace" font-size="8.5" fill="#6b5638" text-anchor="start" font-weight="400" letter-spacing="1.6" >CLEAN UP</text><rect x="740" y="50" width="150" height="70" fill="none" stroke="#2b1f12" stroke-width="0.75"/><text x="815.0" y="82.0" font-family="Space Mono, ui-monospace, monospace" font-size="10.5" fill="#2b1f12" text-anchor="middle" font-weight="700" letter-spacing="1.2" >STOP OR REUSE</text><text x="815.0" y="96.0" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="0.4" >the child process</text><path d="M710 85H738" fill="none" stroke="#2b1f12" stroke-width="0.75" marker-end="url(#lc-arrow)"/><path d="M815.0 120V156H455.0V122" fill="none" stroke="#2b1f12" stroke-width="0.75" marker-end="url(#lc-arrow)" stroke-dasharray="3 3"/><text x="635.0" y="150" font-family="Space Mono, ui-monospace, monospace" font-size="8.5" fill="#6b5638" text-anchor="middle" font-weight="400" letter-spacing="1.6" >REUSE FOR THE NEXT CALL</text></svg>
<figcaption>Fig. 1 &middot; The five steps of a sandboxed call</figcaption>
</figure>
```

## macOS (Seatbelt)

securer writes a Seatbelt profile and starts R with `sandbox-exec`. The
profile denies everything by default, then allows a short list of things R
needs. The child can read R itself, your R library paths, system libraries,
and temp directories, but not your home directory, so files like `~/.ssh`
stay out of reach. It can write only to temp directories. TCP and UDP are
blocked. Unix domain sockets are allowed, because that's how the child talks
to your session.

```{r}
session <- SecureSession$new(sandbox = TRUE)

# Computation works normally:
session$execute("1 + 1")
#> [1] 2

# Network access is blocked:
session$execute('readLines(url("http://example.com"))')
#> Error: Operation not permitted

# Writing outside temp is blocked:
session$execute('writeLines("test", "~/file.txt")')
#> Error: Operation not permitted

session$close()
```

## Linux (bubblewrap)

securer starts R with `bwrap`, which gives the child its own namespaces for
processes, network, users, mounts, host name, and IPC. System libraries and R
are mounted read-only, and `/tmp` is an empty, writable in-memory directory.
Because the child has its own empty network namespace, it has no network.

You need `bwrap` installed (`apt install bubblewrap` on Debian and Ubuntu).
Without it, securer warns you and runs the code without a sandbox.

## Windows

Windows has no sandbox like the ones above. securer gives the child a clean
environment: it keeps only a short list of safe variables, empties
`R_LIBS_USER`, and points `HOME`, `TMPDIR`, `TMP`, and `TEMP` at a private
temp directory. It can also set memory, CPU, and process limits with a Job
Object. It can't restrict files or the network, and it prints a message
saying so the first time you start a session. If you need those
restrictions on Windows, set `SECURER_SANDBOX_MODE=docker-spawn` to run
each session in its own Docker container, or use a Linux VM.

## Resource limits

Resource limits cap what the child process can use. They're set with
`ulimit` and work with or without the sandbox:

```{r}
execute_r("1 + 1", limits = list(cpu = 10, memory = 256 * 1024 * 1024))
```

The limits you can set:

| Name     | Unit    | ulimit flag | Description                 |
|----------|---------|-------------|-----------------------------|
| `cpu`    | seconds | `-t`        | CPU time                    |
| `memory` | bytes   | `-v`        | Virtual address space       |
| `fsize`  | bytes   | `-f`        | Maximum file size           |
| `nproc`  | count   | `-u`        | Maximum processes           |
| `nofile` | count   | `-n`        | Maximum open files          |
| `stack`  | bytes   | `-s`        | Stack size                  |

If you use the sandbox and don't pass `limits`, securer applies these
defaults. Pass `limits = list()` to turn them off:

```{r}
default_limits()
```

## Code pre-validation

`validate_code()` checks code for syntax errors and for calls that look
risky, before the code goes to the child process:

```{r}
# Valid code
validate_code("1 + 1")

# Syntax error
validate_code("if (TRUE {")

# Risky call: a warning, not a block
validate_code("system('ls')")
```

## Checking the sandbox is available

Before you deploy, check that the sandbox tools exist on the machine. If
they're missing, securer runs code without a sandbox, or stops with an
error if `sandbox_strict = TRUE`:

```{r}
if (Sys.info()[["sysname"]] == "Linux") stopifnot(nzchar(Sys.which("bwrap")))
if (Sys.info()[["sysname"]] == "Darwin") stopifnot(file.exists("/usr/bin/sandbox-exec"))
```

## Strict sandbox mode

By default, if the sandbox tools aren't there, securer warns you and runs
the code without a sandbox. In production you probably want it to stop
instead. Set `sandbox_strict = TRUE` along with `sandbox = TRUE` and the
session fails to start if it can't set up the sandbox:

```{r}
# Error if sandbox not available (recommended for production)
session <- SecureSession$new(sandbox = TRUE, sandbox_strict = TRUE)
```

With `sandbox_strict = FALSE`, the default, you get a warning and the
session runs without the OS sandbox. Resource limits, the socket checks, and
the clean environment still apply.

Strict mode doesn't stop a session on Windows, because the clean
environment there counts as the sandbox.

## Architecture

```{=html}
<figure class="hairline">
<svg role="img" aria-label="Sequence between the host and the sandboxed child: start-up, then each call with a tool round trip" viewBox="0 0 900 504" xmlns="http://www.w3.org/2000/svg"><defs>
<marker id="ar-arrow" viewBox="0 0 10 10" refX="9" refY="5" markerWidth="7" markerHeight="7" orient="auto-start-reverse">
<path d="M1 1L9 5L1 9" fill="none" stroke="#2b1f12" stroke-width="1"/></marker>
<pattern id="ar-hatch" width="6" height="6" patternUnits="userSpaceOnUse" patternTransform="rotate(45)">
<line x1="0" y1="0" x2="0" y2="6" stroke="#bf5a36" stroke-width="0.6" opacity="0.55"/></pattern>
</defs><rect x="440" y="20" width="440" height="470" fill="url(#ar-hatch)" opacity="0.35" stroke="none"/><rect x="440" y="20" width="440" height="470" fill="none" stroke="#bf5a36" stroke-width="0.75" stroke-dasharray="4 3"/><text x="872" y="36" font-family="Space Mono, ui-monospace, monospace" font-size="8.5" fill="#bf5a36" text-anchor="end" font-weight="400" letter-spacing="1.6" >OS SANDBOX</text><rect x="140" y="40" width="180" height="36" fill="none" stroke="#2b1f12" stroke-width="0.75"/><text x="230.0" y="62.0" font-family="Space Mono, ui-monospace, monospace" font-size="10.5" fill="#2b1f12" text-anchor="middle" font-weight="700" letter-spacing="1.2" >HOST R SESSION</text><rect x="500" y="40" width="180" height="36" fill="none" stroke="#2b1f12" stroke-width="0.75"/><text x="590.0" y="62.0" font-family="Space Mono, ui-monospace, monospace" font-size="10.5" fill="#2b1f12" text-anchor="middle" font-weight="700" letter-spacing="1.2" >CHILD R PROCESS</text><path d="M230 76V486M590 76V486" stroke="#6b5638" stroke-width="0.6" stroke-dasharray="1 3"/><path d="M30 122H216" stroke="#cdb48a" stroke-width="0.75"/><text x="30" y="116" font-family="Space Mono, ui-monospace, monospace" font-size="8.5" fill="#6b5638" text-anchor="start" font-weight="400" letter-spacing="1.6" >SESSION START &middot; SecureSession$new()</text><path d="M230 158H588" fill="none" stroke="#2b1f12" stroke-width="0.75" marker-end="url(#ar-arrow)"/><text x="410.0" y="151" font-family="Space Mono, ui-monospace, monospace" font-size="9.5" fill="#2b1f12" text-anchor="middle" font-weight="400" letter-spacing="0.8" >START R WITH CALLR</text><text x="604" y="162" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="start" font-weight="400" letter-spacing="0.3" >R starts inside the sandbox</text><path d="M232 198H588" fill="none" stroke="#2b1f12" stroke-width="0.75" marker-start="url(#ar-arrow)" marker-end="url(#ar-arrow)"/><text x="410.0" y="191" font-family="Space Mono, ui-monospace, monospace" font-size="9.5" fill="#2b1f12" text-anchor="middle" font-weight="400" letter-spacing="0.8" >CONNECT THE UNIX SOCKET</text><path d="M230 238H588" fill="none" stroke="#2b1f12" stroke-width="0.75" marker-end="url(#ar-arrow)"/><text x="410.0" y="231" font-family="Space Mono, ui-monospace, monospace" font-size="9.5" fill="#2b1f12" text-anchor="middle" font-weight="400" letter-spacing="0.8" >INJECT THE RUNTIME</text><text x="604" y="242" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="start" font-weight="400" letter-spacing="0.3" >.securer_call_tool() defined</text><path d="M230 278H588" fill="none" stroke="#2b1f12" stroke-width="0.75" marker-end="url(#ar-arrow)"/><text x="410.0" y="271" font-family="Space Mono, ui-monospace, monospace" font-size="9.5" fill="#2b1f12" text-anchor="middle" font-weight="400" letter-spacing="0.8" >INJECT TOOL WRAPPERS</text><text x="604" y="282" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="start" font-weight="400" letter-spacing="0.3" >tool_name() defined</text><path d="M30 308H216" stroke="#cdb48a" stroke-width="0.75"/><text x="30" y="302" font-family="Space Mono, ui-monospace, monospace" font-size="8.5" fill="#6b5638" text-anchor="start" font-weight="400" letter-spacing="1.6" >EACH CALL &middot; $execute(code)</text><path d="M230 344H588" fill="none" stroke="#2b1f12" stroke-width="0.75" marker-end="url(#ar-arrow)"/><text x="410.0" y="337" font-family="Space Mono, ui-monospace, monospace" font-size="9.5" fill="#2b1f12" text-anchor="middle" font-weight="400" letter-spacing="0.8" >THE CODE</text><text x="604" y="348" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="start" font-weight="400" letter-spacing="0.3" >eval(code)</text><path d="M590 384H232" fill="none" stroke="#2b1f12" stroke-width="0.75" marker-end="url(#ar-arrow)"/><text x="410.0" y="377" font-family="Space Mono, ui-monospace, monospace" font-size="9.5" fill="#2b1f12" text-anchor="middle" font-weight="400" letter-spacing="0.8" >{"tool": ...} AS JSON</text><text x="216" y="388" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="end" font-weight="400" letter-spacing="0.3" >runs fn(arg)</text><text x="604" y="388" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="start" font-weight="400" letter-spacing="0.3" >blocks</text><path d="M230 424H588" fill="none" stroke="#2b1f12" stroke-width="0.75" marker-end="url(#ar-arrow)"/><text x="410.0" y="417" font-family="Space Mono, ui-monospace, monospace" font-size="9.5" fill="#2b1f12" text-anchor="middle" font-weight="400" letter-spacing="0.8" >RESULT AS JSON</text><text x="604" y="428" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="start" font-weight="400" letter-spacing="0.3" >carries on</text><path d="M590 464H232" fill="none" stroke="#2b1f12" stroke-width="0.75" marker-end="url(#ar-arrow)"/><text x="410.0" y="457" font-family="Space Mono, ui-monospace, monospace" font-size="9.5" fill="#2b1f12" text-anchor="middle" font-weight="400" letter-spacing="0.8" >PROCESS COMPLETE</text><text x="216" y="468" font-family="Space Mono, ui-monospace, monospace" font-size="9" fill="#6b5638" text-anchor="end" font-weight="400" letter-spacing="0.3" >returns the final value</text></svg>
<figcaption>Fig. 2 &middot; What happens between the two processes, from start-up to a finished call</figcaption>
</figure>
```

securer starts the child through a wrapper script, passed to callr with
`callr::r_session_options(arch = ...)`. The socket lives under `/tmp`,
because macOS limits socket paths to about 104 characters and R's own temp
directory can be longer than that. `vignette("security-model")` goes into
the message format and what each part of the system trusts.

## Why not just callr or Docker?

callr runs code in a separate R process, so a crash there doesn't take down
your session. But it doesn't restrict anything. The child can read your SSH
keys, make HTTP requests, and write anywhere on disk. securer uses callr to
start the child and adds the restrictions.

Docker isolates the filesystem, network, and processes well, but it needs a
running daemon and an image, and containers are slower to start. It also has
no way for the code inside to call back into your R session. You'd have to
build that yourself. securer's docker-spawn backend does use Docker, with a
new container per session, for when you need it, such as on Windows.

By default securer uses the sandbox your operating system already has
(Seatbelt on macOS, bubblewrap on Linux). There's no daemon, sessions start
in under a second, and tool calls, limits, timeouts, and the audit log come
with it.

| Feature                  | callr | Docker | securer |
|--------------------------|-------|--------|---------|
| Process isolation        | Yes   | Yes    | Yes     |
| Filesystem restrictions  | No    | Yes    | Yes     |
| Network restrictions     | No    | Yes    | Yes     |
| Tool-call IPC            | No    | No     | Yes     |
| Resource limits          | No    | Yes    | Yes     |
| Audit logging            | No    | No     | Yes     |
| No daemon required       | Yes   | No     | Yes     |
| Sub-second startup       | Yes   | No     | Yes     |
