---
title: "Integration examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integration examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

These examples put securer inside a Shiny app, a Plumber API, and a batch
script. Each one stands alone. None of the code runs when this page is
built, so copy what you need into your project and adjust it.

### Where securer fits

```
 Shiny / Plumber / Batch script
+------------------------------+
|  Application layer           |     User requests or code snippets
|  (UI, routes, lapply)        |     arrive here.
+---------|--------------------+
          |
          v
+------------------------------+
|  securer                     |     SecureSession or SecureSessionPool
|  - Sandbox (Seatbelt/bwrap) |     manages child R processes with
|  - IPC (Unix domain socket)  |     OS-level isolation.
|  - Resource limits (ulimit)  |
+---------|--------------------+
          |
          v
+------------------------------+
|  Sandboxed R child           |     Code executes here. Tool calls
|  - eval(user code)           |     pause execution and are fulfilled
|  - tool call -> host -> back |     by the host via IPC.
+------------------------------+
```

## Shiny

This small app lets a user type R code and run it in a sandbox. Each
visitor gets their own `SecureSession`, started when they connect and
closed when they leave.

```{r shiny-app}
library(shiny)
library(securer)

ui <- fluidPage(
  titlePanel("Secure R Executor"),
  sidebarLayout(
    sidebarPanel(
      textAreaInput(
        "code", "R Code",
        value = "1 + 1",
        rows = 8
      ),
      actionButton("run", "Run Code")
    ),
    mainPanel(
      verbatimTextOutput("result")
    )
  )
)

server <- function(input, output, session) {
  # One SecureSession per Shiny session

  secure <- SecureSession$new(sandbox = TRUE)

  # Clean up when the user disconnects
  session$onSessionEnded(function() {
    secure$close()
  })

  result_val <- reactiveVal("Enter code and click Run.")

  observeEvent(input$run, {
    result_val(
      tryCatch(
        {
          res <- secure$execute(input$code, timeout = 10)
          paste(capture.output(print(res)), collapse = "\n")
        },
        error = function(e) paste("Error:", conditionMessage(e))
      )
    )
  })

  output$result <- renderText(result_val())
}

shinyApp(ui, server)
```

With many visitors at once, one session each gets expensive. Share a single
`SecureSessionPool` across all of them instead. Each run takes an idle
session from the pool and gives it back when it's done.

## Plumber API

This API takes R code in a POST request and runs it in a pool of sandboxed
sessions. The pool starts with the API and every request shares it.

```{r plumber-setup}
# plumber.R
library(plumber)
library(securer)

# Start the pool once, when the API starts.
# reset_between_uses = TRUE restarts each session after use, so one
# request can't see variables or packages left by another.
pool <- SecureSessionPool$new(
  size = 4,
  sandbox = TRUE,
  reset_between_uses = TRUE
)
```

```{r plumber-endpoint}
#* Execute R code in a sandboxed session
#* @param code Character string of R code
#* @post /execute
function(req, res, code = "") {
  if (nchar(code) == 0) {
    res$status <- 400
    return(list(error = "Missing 'code' parameter"))
  }

  tryCatch(
    {
      result <- pool$execute(code, timeout = 10)
      list(result = result)
    },
    error = function(e) {
      res$status <- 500
      list(error = conditionMessage(e))
    }
  )
}
```

```{r plumber-run}
# Start the API
pr <- plumb("plumber.R")
pr$run(host = "0.0.0.0", port = 8080)
```

Some things to tighten before you expose an API like this:

- Check `nchar(code)` yourself and reject huge requests early.
  `$execute()` already refuses code over `max_code_length` (100,000
  characters by default), but there's no point getting that far.
- Pick a `timeout` that suits your workload. The CPU limit from `ulimit`
  doesn't count time spent waiting, and `timeout` does.
- If you register tools, cap how often one request can call them with
  `max_tool_calls`.
- If error messages go back to people you don't trust, use
  `sanitize_errors = TRUE`. It removes file paths and process IDs.

`SecureSessionPool` doesn't take `max_tool_calls` or `sanitize_errors`. For
those two, manage your own `SecureSession` objects and pass them to `$new()`
and `$execute()`.

## Batch jobs

Here a pool runs a vector of code snippets. `lapply()` goes through them one
at a time, but each snippet runs in a session that's already started, so
you don't pay the startup cost again and again.

```{r batch-snippets}
library(securer)

# Code snippets to evaluate
snippets <- c(
  "mean(1:100)",
  "sqrt(144)",
  "UNDEFINED_VAR + 1",       # will error
  "paste('hello', 'world')",
  "sum(rnorm(1000))"
)
```

```{r batch-execute}
pool <- SecureSessionPool$new(size = 2, sandbox = TRUE)

results <- lapply(snippets, function(code) {
  tryCatch(
    {
      value <- pool$execute(code, timeout = 5)
      list(code = code, result = value, error = NA_character_)
    },
    error = function(e) {
      list(code = code, result = NA, error = conditionMessage(e))
    }
  )
})

pool$close()
```

```{r batch-collect}
# Collect into a data frame for inspection
outcome <- data.frame(
  code  = vapply(results, `[[`, character(1), "code"),
  error = vapply(results, `[[`, character(1), "error"),
  stringsAsFactors = FALSE
)
outcome$result <- lapply(results, `[[`, "result")

outcome[, c("code", "error")]
#>                        code                             error
#> 1             mean(1:100)                               <NA>
#> 2               sqrt(144)                               <NA>
#> 3     UNDEFINED_VAR + 1    object 'UNDEFINED_VAR' not found
#> 4  paste('hello', 'world')                              <NA>
#> 5       sum(rnorm(1000))                               <NA>
```

The `result` column is a list holding each value. Each snippet's error is
caught on its own, so one failure doesn't stop the batch. If you run
snippets in parallel, raise `size` in `SecureSessionPool$new()`. Each
session is a separate R process, so more sessions means more memory.
