---
title: "Quick Start"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
can_eval <- requireNamespace("securer", quietly = TRUE)
```

## Overview

securer runs R code in a sandboxed child process with OS-level isolation.
Code in the child can call named "tools" --- host-side functions that execute
outside the sandbox. Communication happens over a Unix domain socket using JSON
messages. The package is designed for LLM agent systems where generated code
needs access to host-provided capabilities while being prevented from reaching
the network or filesystem.

## Installation

```{r}
pak::pak("ian-flores/securer")
```

Verify the installation:

```{r}
library(securer)
execute_r("1 + 1")
#> [1] 2
```

## Hello world

The simplest way to run code is with `execute_r()`. It creates a session, runs
the code, and tears everything down automatically:

```{r}
library(securer)

execute_r("1 + 1")
#> [1] 2

execute_r("paste('Hello from', R.version.string)")
#> [1] "Hello from R version 4.4.2 (2024-10-31)"
```

By default `execute_r()` enables the OS sandbox (`sandbox = TRUE`). Pass
`sandbox = FALSE` to disable it.

## Defining tools

Tools let sandboxed code call functions on the host. Define them with
`securer_tool()`:

```{r}
add_tool <- securer_tool(
  name = "add",
  description = "Add two numbers",
  fn = function(a, b) a + b,
  args = list(a = "numeric", b = "numeric")
)
add_tool
```

Each tool has four components:

- **name** -- the function name available in the child process
- **description** -- metadata (useful for LLM tool-use prompts)
- **fn** -- the implementation, which runs on the host side
- **args** -- argument names mapped to type strings for validation

### Supported type annotations

| Type string   | Check function     |
|---------------|--------------------|
| `"numeric"`   | `is.numeric()`     |
| `"character"` | `is.character()`   |
| `"logical"`   | `is.logical()`     |
| `"integer"`   | `is.integer()`     |
| `"list"`      | `is.list()`        |
| `"data.frame"`| `is.data.frame()`  |

Type annotations are optional. Arguments without type annotations skip
validation.

## Using tools

Pass tools as a list to `execute_r()`:

```{r}
tools <- list(
  securer_tool("add", "Add two numbers",
    fn = function(a, b) a + b,
    args = list(a = "numeric", b = "numeric")),
  securer_tool("get_weather", "Get weather for a city",
    fn = function(city) list(temp = 72, condition = "sunny"),
    args = list(city = "character"))
)

execute_r('add(2, 3)', tools = tools)
#> [1] 5

execute_r('get_weather("Boston")', tools = tools)
#> $temp
#> [1] 72
#>
#> $condition
#> [1] "sunny"
```

## Next steps

Now that you've seen the basics:

- `vignette("sessions-and-tools")` --- persistent sessions and advanced features
- `vignette("deployment")` --- sandboxing and resource limits
- `vignette("security-model")` --- the full threat model
