---
title: "Quick start"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
can_eval <- requireNamespace("securer", quietly = TRUE)
```

## What securer does

securer runs R code in a separate R process that the operating system keeps
in a sandbox. That code can call "tools", which are functions you define.
The tools run in your own R session, outside the sandbox, and the two
processes pass messages over a Unix domain socket.

It's built for LLM agents. The model writes R code, and the code can use the
tools you give it, but it can't reach the network or your files on its own.

## Installation

```{r}
pak::pak("ian-flores/securer")
```

Verify the installation:

```{r}
library(securer)
execute_r("1 + 1")
#> [1] 2
```

## Hello world

The simplest way to run code is `execute_r()`. It starts a session, runs the
code, and shuts the session down again:

```{r}
library(securer)

execute_r("1 + 1")
#> [1] 2

execute_r("paste('Hello from', R.version.string)")
#> [1] "Hello from R version 4.4.2 (2024-10-31)"
```

`execute_r()` uses the sandbox by default (`sandbox = TRUE`). Pass
`sandbox = FALSE` to turn it off.

## Defining tools

A tool is a function that sandboxed code can call but that runs in your
session. Make one with `securer_tool()`:

```{r}
add_tool <- securer_tool(
  name = "add",
  description = "Add two numbers",
  fn = function(a, b) a + b,
  args = list(a = "numeric", b = "numeric")
)
add_tool
```

A tool has a `name`, which is what the sandboxed code calls it, and a
`description`, which you can show to an LLM so it knows what the tool is
for. `fn` is the function that actually runs, in your session. `args` lists
the argument names and, optionally, the type each one must have.

### Argument types

| Type string   | Check function     |
|---------------|--------------------|
| `"numeric"`   | `is.numeric()`     |
| `"character"` | `is.character()`   |
| `"logical"`   | `is.logical()`     |
| `"integer"`   | `is.integer()`     |
| `"list"`      | `is.list()`        |
| `"data.frame"`| `is.data.frame()`  |

If you leave out an argument's type, securer doesn't check it.

## Using tools

Pass a list of tools to `execute_r()`:

```{r}
tools <- list(
  securer_tool("add", "Add two numbers",
    fn = function(a, b) a + b,
    args = list(a = "numeric", b = "numeric")),
  securer_tool("get_weather", "Get weather for a city",
    fn = function(city) list(temp = 72, condition = "sunny"),
    args = list(city = "character"))
)

execute_r('add(2, 3)', tools = tools)
#> [1] 5

execute_r('get_weather("Boston")', tools = tools)
#> $temp
#> [1] 72
#>
#> $condition
#> [1] "sunny"
```

## Next steps

- `vignette("sessions-and-tools")` covers sessions you keep open, streaming output, and pools.
- `vignette("deployment")` covers the sandboxes and resource limits.
- `vignette("security-model")` covers what securer protects against, and what it doesn't.
